An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-04 EconGrowthSeminar.Rnw
Fixed several small things on the seminar in preparation for the seminar.

\date[Math Seminar] % (optional)
{Mathematics Colloquium\\
Thursday, 4 April 2013
}

@@

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \vspace{\baselineskip}
    \pause
    \item{Energy (relatively) unimportant for classical economists}
    \item{Usually ignored}
    \vspace{\baselineskip}
    \pause
    \item{And yet \dots}
  \end{itemize}
\end{frame}

\begin{frame}{Oil and Recessions}{\citep{Hamilton:2011fo}} %----------------------------
  \begin{center}
    10 of last 11 US recessions preceeded by oil price shocks
    \vspace{\baselineskip}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Set aside debates about growth and its measure}
    \item{All factors and outputs indexed to initial year: $y = Y/Y_0, e = E/E_0$, etc.}
    \vspace{\baselineskip}
    \pause

    \item{Uncritical use of energy-augmented growth models in the literature}
  \end{itemize}
\end{frame}

\begin{frame}{What's Needed?} %----------------------------
  \begin{itemize}
    \item{Go beyond developed economies}
    \pause
    \vspace{\baselineskip}

    \item{Rigorous, statistical evaluation of energy-augmented growth models with $Q$, $X$, and $U$}
    \pause
    \vspace{\baselineskip}

    \item{Comparative analysis of growth models}
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \pause
    \vspace{\baselineskip}

@@

\begin{frame}{Type of Energy ($E$)} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$)}
    \vspace{\baselineskip}  
    \item{Exergy ($X$)}
    \vspace{\baselineskip}
    \item{Useful work ($U$)}
  \end{itemize}
  \vspace{\baselineskip}
  \pause
  \begin{center}Simplification: One type of energy ($Q$) for this presentation\end{center}
\end{frame}

@@

\begin{frame}{LINEX}
    \begin{center} \linexEqn \end{center}
    \vspace{\baselineskip}

@@

\majorSeparator{Data Sources}

@@

<<Coefs_US_SF>>=
sfUSK <- coef(singleFactorModel("US", "K"))
sfUSL <- coef(singleFactorModel("US", "L"))
sfUSQ <- coef(singleFactorModel("US", "Q"))
sfUSX <- coef(singleFactorModel("US", "X"))
sfUSU <- coef(singleFactorModel("US", "U"))
@
\begin{frame}{Single-factor (with $q$) Predictions for $y$} %----------------------------
  \begin{columns}[c]
    \column{2.2in}
      \begin{center}
        \sfEqn
        \vspace{\baselineskip}

@@

\begin{frame}{Comments on Single-factor Model with $q$} %----------------------------
  \begin{itemize}
    \item{Single-factor not typically used for economic modeling: illustrates confidence intervals}
    \vspace{\baselineskip}

    \item{$CI_{\gamma}$ > $CI_{\lambda}$ for all countries\footnote{\tiny as percentage of 0.0 1/year < $\lambda$ < 0.1 1/year and $0 < \gamma < 1$ \normalsize}}
  \end{itemize}
\end{frame}

@@

        \tiny
          \begin{tabular}{|c|c|c|c|c|}\hline
            $e$ & $\lambda$ [1/year] & $\alpha$ & $\beta$ & $\gamma$\\
            \hline
            $-$   & \Sexpr{cdSA["CD", "lambda"]} & \Sexpr{cdSA["CD", "alpha"]} & \Sexpr{cdSA["CD", "beta"]} & \Sexpr{cdSA["CD", "gamma"]}\\
            \hline
            $q$   & \Sexpr{cdqSA["CD", "lambda"]} & \Sexpr{cdqSA["CD", "alpha"]} & \Sexpr{cdqSA["CD", "beta"]} & \Sexpr{cdqSA["CD", "gamma"]}\\
            \hline
            $x$   & \Sexpr{cdxSA["CD", "lambda"]} & \Sexpr{cdxSA["CD", "alpha"]} & \Sexpr{cdxSA["CD", "beta"]} & \Sexpr{cdxSA["CD", "gamma"]}\\
            \hline
          \end{tabular}
        \normalsize
      \end{center}
    \column{2.2in}
<<Results_Graph_SA_CD, fig.width=presentationGraph2ColWidth, fig.height=presentationGraph1RowHeight, fig.pos="H">>=
createCDLatticeGraph("SA", textScaling=textScaleForGraphs, keyXLoc=keyXLoc, keyYLoc=keyYLoc)
@
  \end{columns}
  % Including at bottom of frame so as not to push other material downward in the frame.
  \putFlagUpperRight{SA}{0.055}
\end{frame}

@@

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with $q$} %----------------------------
<<CI_gamma_vs_gamma_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="Q")
@
\end{frame}

\begin{frame}{Implications from Cobb-Douglas with $q$} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Implications from CES} %----------------------------
  \begin{center} \cesEqn \end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Implications from LINEX} %----------------------------
  \begin{center}\linexEqn\end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Summary of Analysis} %----------------------------
  \begin{itemize}
  \item{Including energy ($E$) in Cobb-Douglas and CES production functions appears to be unwarranted}
  \pause
    \begin{itemize}
      \scriptsize
      \item{Energy parameters ($\gamma$, $\delta$) less important than others ($\lambda$, $\alpha$, $\beta$) for nearly all countries}
      \pause
      \item{Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$)}
      \normalsize
    \end{itemize}
  \vspace{\baselineskip}
  \pause
  \item{Including energy ($E$) in LINEX production function is required}
  \pause
    \begin{itemize}
      \scriptsize
      \item{One factor of production: energy}
      \pause
      \item{$a_0$ and $c_t$ are significant for most countries}
      \pause
      \item{Unfortunately, LINEX fits are poorer quality than CD or CES}
      \normalsize
    \end{itemize}
  \item{Uncritical use of energy-augmented growth models is problematic}
  \end{itemize}
\end{frame}

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
  \end{center}
  \vspace{\baselineskip}

\end{frame}

\begin{frame}{The (Tentative) ``Answer''} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}

    Answer: Cobb-Douglas without energy (possibly CES without energy)
  \end{center}
\end{frame}

@@

    \item{Models (CD, or CES, without energy) could be \emph{included within} a model that endogenizes price and cost share}
    \pause
    \item{Still searching for a modeling approach that captures price and cost share dynamics \\ \scriptsize System dynamics modeling? \normalsize}
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Evaluate CES without energy}
    \item{Verify and improve estimates of statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \item{Incorporate all constraints for all models}
    \vspace{\baselineskip}
    \item{Submit paper}
  \end{itemize}  
\end{frame}

\begin{frame}{Acknowledgements} %----------------------------
  \begin{itemize}
    \item{Martin de Wit, Stellenbosch University}
    \item{Randall Pruim, Calvin College}
    \item{Loren L. Heun, Western Michigan University}
    \vspace{\baselineskip}
    \item{Jack and Lois Kuipers Applied Mathematics Endowment\\(Reese and Timmer Summer 2012 funding)}
    \item{Calvin College Alumni Association\\(Reese Fall 2012 funding)}
  \end{itemize}  
\end{frame}

@@

\minorSeparator{CES Model}

>>>>

\date[Math Seminar] % (optional)
{Calvin College Mathematics Colloquium\\
3:30 PM, Thursday, 4 April 2013
}

@@

\begin{frame}{Energy Cost Share is Low} %----------------------------
  \begin{itemize}
    \item{$\sim$10\%}
    \vspace{\baselineskip}
    \pause
    \item{Energy (relatively) unimportant for neoclassical economists}
    \item{Usually ignored}
    \vspace{\baselineskip}
    \pause
    \item{And yet\dots}
  \end{itemize}
\end{frame}

\begin{frame}{Oil and Recessions}{\citep{Hamilton:2011fo}} %----------------------------
  \begin{center}
    10 of last 11 US recessions preceded by oil price shocks
    \vspace{\baselineskip}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output ($Y$) over time}
    \item{Set aside debates about growth and its measure}
    \item{All factors and outputs indexed to initial year: $y = Y/Y_0, e = E/E_0$, etc.}
    \vspace{\baselineskip}
    \pause

    \item{Uncritical use of energy-augmented growth models for developed economies in the literature}
  \end{itemize}
\end{frame}

\begin{frame}{What's Needed?} %----------------------------
  \begin{itemize}
    \item{Move beyond developed economies}
    \pause
    \vspace{\baselineskip}

    \item{Statistically rigorous evaluation of energy types $Q$, $X$, and $U$}
    \pause
    \vspace{\baselineskip}

    \item{Comparative analysis of energy-augmented growth models}
  \end{itemize}
\end{frame}

@@

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which model best describes (statistically) the role of energy in economic growth for different types of economies?
    \pause
    \vspace{\baselineskip}

@@

\begin{frame}{Type of Energy ($E$)} %----------------------------
  \begin{itemize}
    \item{Thermal energy ($Q$)}
    \vspace{\baselineskip}  
    \item{Exergy ($X$)}
    \vspace{\baselineskip}
    \item{Useful work ($U$)}
  \end{itemize}
  \vspace{\baselineskip}
  \pause
  \begin{center} Simplification: One type of energy ($Q$) for this presentation \end{center}
\end{frame}

@@

\begin{frame}{LINEX} %----------------------------
    \begin{center} \linexEqn \end{center}
    \vspace{\baselineskip}

@@

\majorSeparator{Data Sources} %----------------------------

@@

<<Coefs_US_SF>>=
sfUSK <- coef(singleFactorModel("US", "K"))
sfUSL <- coef(singleFactorModel("US", "L"))
sfUSQ <- coef(singleFactorModel("US", "Q"))
sfUSX <- coef(singleFactorModel("US", "X"))
sfUSU <- coef(singleFactorModel("US", "U"))
@
\begin{frame}{Single-factor Predictions for $y$} %----------------------------
  \begin{columns}[c]
    \column{2.2in}
      \begin{center}
        \sfEqn
        \vspace{\baselineskip}

@@

\begin{frame}{Comments on Single-factor Model with $q$} %----------------------------
  \begin{itemize}
    \item{Single-factor model not typically used for economic modeling: illustrates confidence intervals}
    \vspace{\baselineskip}

    \item{$CI_{\gamma}$ > $CI_{\lambda}$ for all countries\footnote{\tiny as percentage of 0.0 yr$^{-1}$ < $\lambda$ < 0.1 yr$^{-1}$ and $0 < \gamma < 1$ \normalsize}}
  \end{itemize}
\end{frame}

@@

        \tiny
          \begin{tabular}{|c|c|c|c|c|}\hline
            $e$ & $\lambda$ [1/year] & $\alpha$ & $\beta$ & $\gamma$\\
            \hline
            $-$   & \Sexpr{pprint(cdSA["CD", "lambda"], digits=4)} & \Sexpr{pprint(cdSA["CD", "alpha"])} & \Sexpr{pprint(cdSA["CD", "beta"])} & \Sexpr{pprint(cdSA["CD", "gamma"])}\\
            \hline
            $q$   & \Sexpr{pprint(cdqSA["CD", "lambda"], digits=4)} & \Sexpr{pprint(cdqSA["CD", "alpha"])} & \Sexpr{pprint(cdqSA["CD", "beta"])} & \Sexpr{pprint(cdqSA["CD", "gamma"])}\\
            \hline
            $x$   & \Sexpr{pprint(cdxSA["CD", "lambda"], digits=4)} & \Sexpr{pprint(cdxSA["CD", "alpha"])} & \Sexpr{pprint(cdxSA["CD", "beta"])} & \Sexpr{pprint(cdxSA["CD", "gamma"])}\\
            \hline
          \end{tabular}
        \normalsize
      \end{center}
    \column{2.2in}
<<Results_Graph_SA_CD, fig.width=presentationGraph2ColWidth, fig.height=presentationGraph1RowHeight, fig.pos="H">>=
createCDLatticeGraph("SA", textScaling=textScaleForGraphs, keyXLoc=keyXLoc, keyYLoc=keyYLoc)
@
  \end{columns}
  % Including at bottom of frame so as not to push other material downward in the frame.
  \putFlagUpperRight{SA}{0.055}
\end{frame}

@@

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with $q$} %----------------------------
<<CI_gamma_vs_gamma_Graph_Q, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="Q")
@
\end{frame}

\begin{frame}{Summary of Cobb-Douglas with $q$} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Summary of CES with $q$} %----------------------------
  \begin{center} \cesEqn \end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Summary of LINEX with $q$} %----------------------------
  \begin{center}\linexEqn\end{center}
  \vspace{\baselineskip}

@@

\begin{frame}{Summary of Analysis} %----------------------------
  \begin{itemize}
  \item{Augmenting Cobb-Douglas and CES production functions with energy ($E$) appears to be unwarranted}
  \pause
    \begin{itemize}
      \scriptsize
      \item{Energy parameters ($\gamma$, $\delta$) less important than others for nearly all countries}
      \pause
      \item{Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$)}
      \normalsize
    \end{itemize}
  \pause
  \item{Uncritical use of energy-augmented CD and CES growth models is problematic}
  \vspace{\baselineskip}
  \pause
  \item{Including energy ($E$) in LINEX production function is required}
  \pause
    \begin{itemize}
      \scriptsize
      \item{One factor of production: energy}
      \pause
      \item{$a_0$ and $c_t$ are significant for most countries}
      \pause
      \item{LINEX fits are poorer quality than CD or CES}
      \normalsize
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which model best describes (statistically) the role of energy in economic growth for different types of economies?
  \end{center}
\end{frame}

\begin{frame}{The (Tentative) ``Answer''} %----------------------------
  \begin{center}
    Cobb-Douglas without energy\\(possibly CES without energy)
  \end{center}
\end{frame}

@@

    \item{Models (CD, or CES, without energy) could be \emph{included within} a system dynamics model that endogenizes price and cost share}
    \pause
    \item{System dynamics modeling}
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Evaluate CES without energy}
    \vspace{\baselineskip}
    \item{Verify and improve estimates of statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \item{Incorporate all constraints for all models}
    \vspace{\baselineskip}
    \item{Submit paper}
    \vspace{\baselineskip}
    \item{Develop system dynamics model}
  \end{itemize}  
\end{frame}

\begin{frame}{Acknowledgements} %----------------------------
  \begin{itemize}
    \item{Martin de Wit, Stellenbosch University}
    \item{Loren L. Heun, Western Michigan University}
    \item{Randall Pruim, Calvin College}
    \vspace{\baselineskip}
    \item{Jack and Lois Kuipers Applied Mathematics Endowment\\(Reese and Timmer Summer 2012 funding)}
    \item{Calvin College Alumni Association\\(Reese Fall 2012 funding)}
  \end{itemize}  
\end{frame}

@@

\begin{frame}{$CI_{\gamma}$ and $\gamma$ in Cobb-Douglas} %----------------------------
  \begin{center}
    \cdeEqn\\
    \includegraphics[height=2.5in]{figure_other/CI-gamma-4.pdf}
  \end{center}
\end{frame}

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with $x$} %----------------------------
<<CI_gamma_vs_gamma_Graph_X, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="X")
@
\end{frame}

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with $u$} %----------------------------
<<CI_gamma_vs_gamma_Graph_U, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="U")
@
\end{frame}

\minorSeparator{CES Model} %----------------------------
